### 2020-11-12 docs/manuscript.tex
add 'Identifying HLA-escape mutations' section

One way how the human immune system is able to recognize intracellular viral infections
is through the human leucocyte antigen system:
In cells with active protein biosynthesis, proteins are continuously synthesized
and also degraded by a process called proteasomal degradation, which cleaves
proteins into linear peptides of varying length.
A small subset of these peptides is presented on the cell surface via
receptors called MHC class I. The genomic region encoding for MHC I is known to be
highly polymorphic, with more than 20000 different HLA alleles described today.
The resulting gene products differ in their binding properties, which means that
cells from different individuals present a highly diverse set of peptides on their
surface.
Cytotoxic T cells are selected during maturation to only weakly bind to 
peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, 
but might be able to strongly bind to complexes of MHC I with peptides which are 
generated through proteosomal degradation of a viral protein.
Upon activation, T cells induce cytolytic activity and recruit other immune cells.

@@

The evolutionary events are complex and occur not only on the level of individuals, 
where a virus adapts to specific features of the host, but also on the population level,
because HLA alleles differ in their frequency across geographic region, as they are inherited
according to standard Mendelian rules.

Upon transmission, HLA escape mutations typically quickly revert to their wild type
as they usually reduce viral replicative capacity (if a mutation would increase
viral replicative capacity regardless of the presence of a given HLA allele, it would
probably already be the wild type). Kawashima et al
describe an escape mutation that is selected by HLA allele B\*53,
does not strongly affect viral replicative capacity and therefore slowly enriches
over time in Japan, where B\*53 commonly occurs.
How quickly a given escape mutation is selected upon transmission in a host depends on the
magnitude of the reduction in viral replicative capacity, on the strength of selection
pressure and also on the genetic background, e.g. some escape mutations require compensatory 
mutations which partly attenuate the negative impact on viral replicative capacity.

Studying HLA escape therefore provides the unique opportunity to gain insight into
viral evolution, on the host level but also on the population level.
Unfortunately, identifying HLA escape mutations is difficult in practice.

@@

>>>>

One way how the human immune system is able to recognize intracellular viral infections
is through the human leucocyte antigen system:
In cells with active protein biosynthesis, proteins are continuously synthesized
and also degraded by a process called proteasomal degradation, which cleaves
proteins into linear peptides of varying length.
A small subset of these peptides is presented on the cell surface via
receptors called MHC class I. The genomic region encoding for MHC I is known to be
highly polymorphic, with more than 20000 different HLA alleles described today.
The resulting gene products differ in their binding properties, which means that
cells from different individuals present a highly diverse set of peptides on their
surface.
Cytotoxic T cells are selected during maturation to only weakly bind to 
peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, 
but might be able to strongly bind to complexes of MHC I with peptides which are 
generated from of a viral protein.
Upon activation, T cells induce cytolytic activity and recruit other immune cells.

@@

The evolutionary events are complex and occur not only on the level of individuals, 
where a virus adapts to specific features of the host, but also on the population level,
because HLA alleles differ in their frequency across geographic regions, as they are inherited
according to standard Mendelian rules.

Upon transmission, HLA escape mutations typically quickly revert to their wild type
as they usually reduce viral replicative capacity (if a mutation would increase
viral replicative capacity regardless of the presence of a given HLA allele, it would
probably already be the wild type). Kawashima et al.
describe an escape mutation that is selected by HLA allele B\*53,
does not strongly affect viral replicative capacity and therefore slowly enriches
over time in Japan, where B\*53 commonly occurs.
How quickly a given escape mutation is selected upon transmission in a host depends on the
magnitude of the reduction in viral replicative capacity, on the strength of selection
pressure and also on the genetic background, e.g. some escape mutations require compensatory 
mutations which partly attenuate the negative impact on viral replicative capacity.

Studying HLA escape therefore provides an unique opportunity to gain insight into
viral evolution, on the host level but also on the population level.
Unfortunately, identifying HLA escape mutations is difficult in practice.

@@

There are several experimental methods available to study HLA escape:
Recombinant MHC-I molecules can be used in binding assays: Upon complex formation
with a peptide, a change in conformation can be detected with conformation-specific
antibodies. This method is relatively fast, but only measures binding affinity of a
peptide to MHC-I and does not account for antigen processing or immunodominance, 
which describes the observation that a peptide may be presented via MHC-I on the cell surface but does
not induce an immune response.
An experimental setup that resembles the conditions in-vitro more closely but is also
more time-consuming is to measure
CD8+ T cell responses instead. This is usually done by stimulating PBMCs with prototype
and variant peptides and measuring the secretion of IFN-gamma by intracellular cytokine
staining and fluorescence-assisted cell sorting.
To analyse CD8+ T cell responses against endogenously processed antigens it is necessary
to generate cell-lines stably expressing the antigen in question and adding antigen-specific
CD8+ T cells. This method scales poorly as it requires transfection of cell lines and
antigen-specific expansion of CD8+ T cells.
